LEMONADE: Leveraging Modern Automated Deduction for Program Synthesis\\ \vspace{0.5cm} \large{WWTF ICT 2022}

### 2022-04-04 WWTF/main.tex
asdf

The main goal of this project is to connect the classical approach to deductive synthesis --  proofs-as-programs~\cite{girard1989proofs} --, with modern automated theorem provers, in particular first-order solvers (including SMT solvers), which have made impressive progress over the recent years (as evidenced by yearly competitions\footnote{\url{http://www.tptp.org/CASC/}}).

We believe this approach offers the following advantages:
1) First-order solvers are particularly suitable to solving \emph{quantified} formulas as needed for synthesis.
2) Proofs-as-programs represents a longstanding approach to program extraction.
That is to say, much groundwork for program extraction from proofs has been laid, we just need to adapt and extend this theory to the specifics of modern automated theorem provers.
3) We are able to take advantage of the highly sophisticated and optimized proof search strategies of first-order solvers instead of needing to define and implement \emph{ad-hoc} search strategies.
%Moreover, we will profit from any future advancements in automated theorem proving.

We believe the proposed research is very timely, because improved hardware and the progress in automated theorem proving have finally made automated proof search feasible.
Thus the proposed research promises to fill a longstanding gap in the state-of-the-art.

We will advance the above main goal by two work packages (WPs):

@@

The final goal of our project is to leverage state-of-the-art first-order automated
theorem provers, in particular Vampire, for program synthesis.
This is motivated by the recent and prospective advancements in their capabilities.
We will lay the necessary theoretical foundations as well as implement a prototype system to leverage state-of-the-art classical theorem provers for program extraction.
At the current time no such system exists.

@@

We now give details on (WP1).
We will consider two approaches from the literature:

1) Parigot's $\lambda\mu$-calculus~\cite{Parigot1} gives a direct translation of classical proofs with datatypes and is essentially ready for implementation.
Unfortunately, we loose the realizability of $\exists$ and $\vee$, since non-constructive tautologies like $\neg\forall xA(x)\to \exists x\neg A(x)$ or $A\vee \neg A$ are classically provable.
As discussed in~\cite{practical}, in practice it is sometimes possible to eliminate the non-constructive reasoning steps from proofs.
We aim to establish stronger theoretical guarantees on when this is possible but also evaluate this approach when realizability is not ensured by a theorem.

2) Sometimes classical validity guarantees intuitionistic validity, most famously for $\Pi_2$-formulas in Arithmetic~\cite{Friedman}. There have been many variations of this result, most notably by Schwichtenberg's group~\cite{schwichtenberg}.
We will collect, adapt and extend these results with regard to our concrete setting.
This is particularly promising when applied to SMT, which can be interpreted as a restricted subproblem of the full first-order case~\cite{reynolds2015counterexample}, or when considering altered specifications (see next point).
%We expect that in practice these approaches will be sufficient in many situations.

Third, we investigate a new direction:
We aim for a translation between provers rather than proofs.
E.g., we study under what conditions classical proofs of a modified input formula allow to extract intuitionistic proofs of the actual input.
A motivating example is actually the reverse direction:
given an intuitionistic prover one can establish classical provability via double negation translation.
%A first approximation has been outlined in~\cite{RDNT}.

Using the results of (WP1) we will implement a synthesis system on top of Vampire (WP2), targeting a suitable functional programming language.
We will experiment with various benchmark problems (e.g., median computation, sorting algorithms, arithmetic functions such as sqrt) and compare our results with the SyGuS-competition.
We will use ideas from syntax-guided synthesis (sketched in 1) in order to restrict and guide proof search:
We will experiment with different formulations of arithmetic axioms, induction axioms and redundant axioms.
We will further examine if proof search is possible when only input/output pairs are given.

@@

>>>>

The main goal of this project is to connect the classical approach to deductive synthesis --- proofs-as-programs~\cite{girard1989proofs} ---, with modern automated theorem provers, in particular first-order solvers (including SMT solvers), which have made impressive progress over the recent years (as evidenced by yearly competitions).
%\footnote{\url{http://www.tptp.org/CASC/}})

We believe that this approach offers the following advantages:
1) First-order solvers are particularly suitable to solving \emph{quantified} formulas as needed for synthesis.
2) Proofs-as-programs represents a longstanding approach to program extraction.
That is to say, much groundwork for program extraction from proofs has been laid, we just need to adapt and extend this theory to the specifics of modern automated theorem provers.
3) We are able to take advantage of the highly sophisticated and optimized proof search strategies of first-order solvers instead of needing to define and implement \emph{ad-hoc} search strategies.
%Moreover, we will profit from any future advancements in automated theorem proving.

We believe the proposed research is very timely, because improved hardware and the progress in automated theorem proving have finally made automated proof search feasible.
We will advance the above main goal by two work packages (WPs):

@@

The final goal of our project is to leverage state-of-the-art first-order automated
theorem provers, in particular Vampire, for program synthesis.
This is motivated by the recent and prospective advancements in their capabilities.
We will lay the necessary theoretical foundations as well as implement a prototype system to leverage state-of-the-art classical theorem provers for program extraction.
At the current time no such system exists.
Thus the proposed research promises to fill a longstanding gap in the state-of-the-art.

@@

We now give details on (WP1):
%We will consider two approaches from the literature:

1) Parigot's $\lambda\mu$-calculus~\cite{Parigot1} gives a direct translation of classical proofs with datatypes and is essentially ready for implementation.
Unfortunately, we loose the realizability of $\exists$ and $\vee$, since non-constructive tautologies like $\neg\forall xA(x)\to \exists x\neg A(x)$ or $A\vee \neg A$ are classically provable.
In practice it is sometimes possible to eliminate the non-constructive reasoning steps from proofs~\cite{practical}.
We aim to establish stronger theoretical guarantees on when this is possible but also evaluate this approach when realizability is not ensured by a theorem.

2) Sometimes classical validity guarantees intuitionistic validity, most famously for $\Pi_2$-formulas in Arithmetic~\cite{Friedman}. There have been many variations of this result, most notably by Schwichtenberg's group~\cite{schwichtenberg}.
We will collect, adapt and extend these results with regard to our concrete setting.
This is particularly promising when applied to SMT, interpreted as a restricted subproblem of the full first-order case~\cite{reynolds2015counterexample}, or when considering altered specifications (see next point).
%We expect that in practice these approaches will be sufficient in many situations.

3) We investigate a new direction:
We aim for a translation between provers rather than proofs.
E.g., we study under what conditions classical proofs of a modified input formula allow to extract intuitionistic proofs of the actual input.
A motivating example is actually the reverse direction:
given an intuitionistic prover one can establish classical provability via double negation translation.
%A first approximation has been outlined in~\cite{RDNT}.

Using the results of (WP1) we will implement a synthesis system on top of Vampire (WP2), targeting a suitable functional programming language.
We will experiment with various benchmark problems %(e.g., median computation, sorting algorithms, arithmetic functions such as sqrt) 
and compare our results with the SyGuS-competition.
We will use ideas from syntax-guided synthesis in order to restrict and guide proof search:
We will experiment with different formulations of arithmetic axioms, induction axioms and redundant axioms.
We will further examine if proof search is possible when only input/output pairs are given.

@@

%[1] Ulrich Berger, Wilfried Buchholz, and Helmut Schwichtenberg. Refined
%program extraction from classical proofs. 2002.
%
%[2] Dana Fisman, Rishabh Singh, and Armando Solar-Lezama. Special issue
%on syntax-guided synthesis preface. 2022.
%
%[3] Harvey Friedman. Classically and intuitionistically provably recursive
%functions. 1978.
%
%[4] Jean-Yves Girard, Paul Taylor, and Yves Lafont. Proofs and types. 1989.
%
%[5] Timothy G Griffin. A formulae-as-type notion of control. 1989.
%
%[6] Ulrich Kohlenbach. The Friedman A-translation. 2008.
%
%[7] Yevgeniy Makarov. Practical program extraction from classical proofs.  2006.
%
%[8] Zohar Manna and Richard Waldinger. A deductive approach to program
%synthesis. 1979.
%
%[9] Michel Parigot. λμ-calculus: an algorithmic interpretation of classical
%natural deduction. 1992.
%
%[10] Andrew Reynolds, Morgan Deters, Viktor Kuncak, Cesare Tinelli, and
%Clark Barrett. Counterexample-guided quantifier instantiation for synthesis
%in SMT. 2015.
